\LARGE \bf GPU Acceleration of Fixed Complexity Sphere Decoder for Large MIMO Uplink Systems

### 2014-11-23 FCSD_GPU_tex/conference draft.tex
conference draft 2

% draft mode.
\documentclass[a4paper, 10pt, conference]{IEEEconf}
%\documentclass[12pt,a4paper,final]{article}
%\IEEEoverridecommandlockouts
%\overrideIEEEmargins
%\documentclass[conference]{IEEEtrans}
% Add the compsoc option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,ft.tex

% manually specify the path to it like:
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{caption}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{epstopdf}
% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

@@

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Do not use the stfloats baselinefloat ability as IEEE does not allow
% \baselineskip to stretch. Authors submitting work to the IEEE should note
% that IEEE rarely uses double column equations and that authors should try
% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
% packages (also by Sigitas Tolusis) as IEEE does not format its papers in
% such ways.
% Do not attempt to use stfloats with fixltx2e as they are incompatible.
% Instead, use Morten Hogholm'a dblfloatfix which combines the features
% of both fixltx2e and stfloats:
%
 \usepackage{dblfloatfix}
% The latest version can be found at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/dblfloatfix/

@@

\begin{document}
% make the title area
%\begin{spacing}{2.5}
\date{}
\maketitle
\thispagestyle{empty}
\pagestyle{empty}
% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
Large MIMO constitutes a principal techniques for next generation wireless communication systems. Detection algorithms for large MIMO tend to require long simulation times on conventional computer systems, in order to reduce the simulation times for decoders of large MIMO systems, we propose to use General Purpose Graphic Processing Units (GPGPU). In this paper Fixed Complexity Sphere Decoder (FCSD) is implemented for the large scale MIMO system based on the dominant framework of GPGPU-Computed Unified Device Architecture (CUDA). The proposed CUDA-FCSD can achieve considerable speedup over conventional CPU implementation while reaching the same Bit Error Rate (BER) performance. Furthermore the proposed algorithm demonstrates great potential to accelerate large MIMO systems.\\
\textbf{ Keywords: Large MIMO, FCSD, GPGPU, CUDA}
\end{abstract}

@@

\section{Introduction}
The present challenge in the wireless communication area is the sharp contrast between the increasing demand of mobile network speed for the transmission of rich media content with better QoS (Quality of Service), and the shortage of radio frequency spectrum. Multiple Input Multiple Output (MIMO) is thought to be a very promising technology that can help to solve this problem owing to several advantages it can offer. The core idea of MIMO is to use multiple antennas that at the transmitting and receiving sides, that provide parallel data pipes in space resulting in spatial multiplexing gains\cite{oestges2010mimo}. For example the LTE standard supports up to 8 antennas at the base station\cite{dahlman20103g}.

Large scale MIMO (also named massive MIMO) systems employ hundreds of low-power low-price antennas collocated at the base station and serving several tens of terminals at the same time\cite{rusek2013scaling}. This technology originated from multiuser MIMO (MU-MIMO) systems\cite{larsson2013massive}. Large scale MIMO will achieve the full potential of conventional MIMO systems (spatial diversity gain and spatial multiplexing gain), as well as provide additional reduction of latency and improvement of system robustness. Detector is one of the key component of MIMO systems,
 a well known detector for MIMO system is the Sphere Decoder (SD) that perform as well as a Maximum Likelihood detector (MLD) with a dramatic complexity reduction\cite{viterbo1999universal}\cite{hassibi2005sphere}. There are two major problems with the traditional SD decoder: complexity depending on Signal to Noise Ratio (SNR), and sequential nature, which are contradictory with the requirement of a constant data processing throughput and parallel implementation for practical system. The Fixed Complexity Sphere Decoder (FCSD) overcomes these two disadvantages of the SD decoder by searching the solution along a fixed number of independent paths\cite{barbero2008fixing}.

General purpose Graphic processing Units Computing (GPGPU) has become a common trend in industry and research in areas that need massive computations, such as computer vision\cite{fung2008using}, computer games\cite{blewitt2013applicability}, signal processing\cite{van2011accelerating}\cite{6671435} and finance\cite{grauer2013accelerating}. Compared to CPU, the GPU can naturally cope with computational intensive tasks, since most of the hardware resource are allocated for data processing and parallel computation, rather than caching and flow control. In recent years GPU computing has provided significant advantages in data processing speeds over CPU computations\cite{nvidia2008programming}. In 2011, the NVIDIA GPUs have become the fastest supercomputer in the world\cite{cook2013cuda},  Computed unified Device Architecture (CUDA) is a programming model for GPGPU that was proposed by NIVIDIA and has become the dominant framework for GPGPU programming. It is based on popular programming languages(C, C++, Java, Python, Fortran, Direct Compute), such that the application developers in different areas do not need to learn a new programming language or get to know the hardware details. CUDA employs the Single Instruction Multiple Threads (SIMT) programming model, enables a GPU to implement parallel data processing, and has become a parallel computing platform for wide community of different applications.

@@

The rest of the paper is organized as follows. Section \ref{system} describes the MIMO systems model. Section \ref{programming model} introduces CUDA, GPU architecture and the programming model. In section \ref{GPUFCSD}, we present details of the implementation of CUDA-FCSD. In section \ref{simulation}, the speedup and bit error rate (BER) performance of CUDA-FCSD are presented and analyzed. Finally conclusions are drawn in section \ref{conclusion}.    
% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.

@@

\section{CUDA GPU Architecture and  Programming Model} \label{programming model}
\subsection{Memory Hierarchy}  
CUDA programming model is a C programming extension, with special defined syntax and structure. The serial code is executed on the host (CPU) side, and massive parallel data processing is performed on the device (GPU) side, this is defined as a heterogeneous programming model. A thread is the basic programming unit of CUDA, a block of cooperative threads form a thread block, in one block all the threads can share one pitch of on chip memory (shared memory), all the threads executed in one block can synchronize their action meaning the threads are permitted to execute the next instruction only when all the threads have reached the same instruction node. Blocks can be grouped into a one-dimensional, two-dimensional, or three-dimensional grid. Similarly, The threads can also be organized into one-dimensional, two-dimensional or three-dimensional blocks. The organization of the block and grid shape and dimension is determined by the application, Figure \ref{figure1}  simply illustrates the three dimensional thread hierarchy of CUDA.
\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth]{CUDA_programming_hierachy.eps}
\caption{CUDA 3D thread hierachy\\ $5\times 5\times 5$ grid and each block contains $5\times 5\times 5$ threads\\}
\label{figure1}
\end{figure}

\subsection{Single Instruction Multiple Thread (SIMT) Model} 
Threads are scheduled, managed, created and executed in a group of 32 threads called warp, the programs executes on the device side are defined as $\mathit{kernel}$ function, when a $\mathit{kernel}$ function is invoked, the thread blocks are partitioned into warps, the threads in one warps executes the same instruction simultaneously, the different warps are scheduled by warp scheduler, when one warp is waiting for data loading the other warps begin to execute, this is called Single Instruction Multiple Thread (SIMT) model. This programming architecture is askin to Single Instruction Multiple Data (SIMD), however unlike the SIMD vectoring machines, SIMT can branch the behavior of each threads to achieve thread-level parallelism. The GPU use massive number of threads to hide the long latency of global memory instead of using large cache as CPU. In a word CPU cores are designed to minimize latency for each thread at a time, whereas GPUs are designed to handle a large number of concurrent, lightweight threads in order to maximize throughput.
\subsection{Mircroarchitecture of GPU} 
In Figure \ref{figure2} we use the GPU-GeForce GTX 760 as an example to show the microarchitecture of  GPU. The GPU works as a coprocessor system, consisting of Stream Multiprocessors (SM). The GeForce GTX 760 has 6 SM that perform the multiply-add arithmetic operation. Furthermore, each SM has the special functional units(SFU) that can perform more complex arithmetics such as trigonometric functions and reciprocal square root. The shared memory, also called on-chip memory located on SM and has low latency and limited capacity. Global memory, which is off-chip memory of the GPU has long latency and large capacity. Another two types of the memory, texture memory and constant memory, are located off-chip, but cached. Therefore the loading speed of the data located in these two type of memory space is much faster than that from global memory. Constant memory is read only and with latency approximately as small as the registers. Registers are allocated dynamically and privately to threads, the access speed of registers is the faster than all the other four types of memories. The variables invoked in $\mathit{kernel}$ functions can be stored in either of the these five types of memories.

Because of the resource limitation of SM the total number of threads can be defined in one block is limited. On current CUDA supported GPUs, one block can contains up to 1024 threads, and all the blocks are executed on SM. Thus the total number of threads ($T_{total}$) we can make use of is $T_{total}=N_{SM}\times N_{threads/SM} $, $N_{SM}$ and $N_{threads/SM}$ denote the number of SM and number of threads per SM. For example the GeForce GTX 760 allows 2048 threads in one SM, thus the total number of threads that can work in parallel is $2048\times 6=12288$.
\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth]{High_view_of_CUDA_GPU_microarchitecture.eps}
\caption{High View of CUDA CPU Microarchitecture}
\label{figure2}
\end{figure}   
% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals/conferences use top floats
% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
% command of the stfloats package.
\section{GPU Based Acceleration of FCSD}\label{GPUFCSD}
\subsection{MIMO System Model}\label{system}
% no \IEEEPARstart
%This demo file is intended to serve as a ``starter file''
%for IEEE conference papers produced under \LaTeX\ using
%IEEEtran.cls version 1.8 and later.
% You must have at least 2 lines in the paragraph with the drop letter
% (should never be an issue)
%I wish you the best of success.
%\hfill mds
%\hfill December 27, 2012

We consider a complex uncoded spatial multiplexing MIMO system with $N_r$ receive and $N_t$ transmit antennas, $N_{r}\geq N_{t}$, over a flat fading channel, corresponding to the discrete time model, the $\mathbb{y}\in\mathbb{C}^{N_{r}\times 1}$ received symbol vecot can be written as:
\begin{equation}
\mathbf{y}=\mathbf{H}\mathbf{s}+\mathbf{n}   \label{formula 1}
\end{equation}
$\mathbf{s}\in \mathbb{C}^{N_{t}\times 1}$ is the transmitted symbol vector, with components that are mutually independent and taken from the signal constellation $\mathbb{O}$ (4-QAM, 16-QAM, 64-QAM), of size $M$. All the possible transmitted symbol vectors forms a $N_{t}$ complex signal vector space $\mathbb{O}^{N_{t}}$. $\mathbb{E}[\mathbf{s}\mathbf{s}^{H}]=\mathbf{I}_{N_t}E_{s}$, where $E_{s}$ denotes the symbol average energy, and $\mathbb{E}[\cdot]$ denotes the expectation operation. $\mathbf{H}\in \mathbb{C}^{N_{r}\times N_{t}}$ denotes the Rayleigh fading channel propagation matrix with independent identity distributed(i.i.d) circularly symmetric complex Gaussian zero mean components of unit variance, $\mathbf{n}\in \mathbb{C}^{N_{r}\times 1}$ is the additive white Gaussian noise (AWGN) samples with zero mean components and $\mathbb{E}[\mathbf{n}\mathbf{n}^{H}]=\mathbf{I}_{N_{r}}N_{0}$, $N_{0}$ denotes the average noise power spectrum density and $\frac{E_{s}}{N_{0}}$ denotes the signal to noise ratio (SNR).

Assume we have the perfect channel state information (CSI), meaning that $ \mathbf{H}$ is known at the receiver, as well as the SNR. The task of MIMO decoder is to recover $\mathbf{s}$ based on $\mathbf{y}$ and $\mathbf{H}$.
\subsection{Fixed Complexity Sphere Decoder}
From (\ref{formula 1}), the maximum likelihood detector (MLD) for a MIMO system is specified by:
\begin{equation}
\mathbf{s}_{ML}=arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{y}-\mathbf{H}\mathbf{s}||^{2} \label{formula 2}
\end{equation}
We consider a MIMO system with  $N_{t}=N_{r}$. Performing the QR decomposition to the channel propagation matrix
\begin{equation}
 \mathbf{H}=\mathbf{Q}\mathbf{R}  \label{QR}
\end{equation}
where $\mathbf{Q}\in \mathbf{C}^{N_{t}\times N_{t}}$ is an unitary matrix and $\mathbf{R}\in \mathbf{C}^{N_{t}\times N_{t}}$, is an upper triangular matrix, the basing on \cite{golub2012matrix}, we can write (\ref{formula 2}) as:
\begin{eqnarray}
\nonumber
\mathbf{s}_{ML}&=&arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{Q}^{H}\mathbf{y}-\mathbf{Q}^{H}\mathbf{H}\mathbf{s}||^{2}\\
&=& arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{Q}^{H}\mathbf{y}-\mathbf{R}\mathbf{s}||^{2} \label{formula 3}
\end{eqnarray}
Consider the unconstrained estimation of $\mathbf{s}$, $\mathbf{\hat{s}}=\mathbf{G}\mathbf{y}$, $\mathbf{G}=(\mathbf{H}^{H}\mathbf{H})^{-1}\mathbf{H}^{H}$ based on (\ref{QR}), we have  
\begin{equation}
\mathbf{\hat{s}}=\mathbf{G}\mathbf{y}=(\mathbf{R}^{H}\mathbf{R})^{-1}\mathbf{R}^{H}\mathbf{Q}^{H}\mathbf{y}
=\mathbf{R}^{-1}\mathbf{Q}^{H}\mathbf{y}
\label{unconstrained estimation}
\end{equation} 
Therefore $\mathbf{Q}^{H}\mathbf{y}=\mathbf{R}\mathbf{\hat{s}}$, and (\ref{formula 3}) can be written as
\begin{equation}
\mathbf{s}_{ML}=arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{R}(\mathbf{\hat{s}}-\mathbf{s})||^{2} \label{formula 4}
\end{equation}

Just like SD the FCSD also performs depth first searching, at the first $\rho$ node levels (\cite{barbero2008fixing} shows when $N_{r}=N_{t}$ the FCSD can achieve the same diversity as the MLD if $\rho\geq \sqrt{N_{t}}-1$) the FCSD search all the possible signal symbols in constellation $\mathbb{O}^{N_{t}}$, which is called Full Expansion(FE) stage, therefore for the symbol nodes $\widetilde{s}_{N_{t}} \widetilde{s}_{N_{t}-1}\dots \widetilde{s}_{N_{t}-\rho+1}$, the FCSD has $M$ branches expansion at each symbol node,  as shown in Figure \ref{figure4}, there are $\rho=2$ FE node levels and each node level has $M=4$ branches. According to (\ref{formula 4}), we have: 
\begin{equation}
\hat{s}^{F}_{i}=arg\min_{s\in \mathbb{O}^{N_{t}}}\{\sum_{i=1}^{N_{t}}|r_{ii}(\hat{s}_{i}-s_{i})+\sum_{j=i+1}^{N_{t}}r_{ij}(\hat{s}_{j}-s_{j})|^{2}\}  \label{formula 5}
\end{equation} 
$i\in [1,2\dots N_{t}]$, where $r_{ij}$ denotes the component at the $i$th row and $j$th column of the upper triangular matrix $\mathbf{R}$,  based on (\ref{formula 5}) at the remaining $N_{t}-\rho$ levels of symbols, the FCSD performs decision feed back, therefore there are only $1$ branch expansion for each symbol node at this stage, this stage is called Single Expansion (SE). 
Thus the total number of the branches is fixed (For example in Figure \ref{figure4} the total branch number is $4^{2}$), the FCSD works as a constant number of multiple path tree searching algorithm.
\begin{figure}[htb]
\includegraphics[width=0.5\textwidth]{FCSD_tree_searching.eps}
\caption{Tree searching of 4-QAM FCSD for 8x8 MIMO system}
\label{figure4}
\end{figure}
In conclusion the FCSD symbol vector candidate $s^{F}_{i}=[s_{1},s_{2}\dots s_{N_{t}}]$ can be expressed as:
\begin{displaymath}
s^{F}_{i}=\left\lbrace\begin{array}{c}
\Omega\in  \mathbb{O}^{\rho}      \quad i=N_{t},N_{t}-1,\dots N_{t}-\rho+1\\

>>>>

% draft mode.
\documentclass[a4paper, 10pt, conference]{ieeeconf}
%\documentclass[letterpaper, 10pt, conference]{ieeeconf}
%\documentclass[12pt,a4paper,final]{article}
%\IEEEoverridecommandlockouts
%\overrideIEEEmargins
%\documentclass[conference]{IEEEtrans}
% Add the compsoc option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,ft.tex

% manually specify the path to it like:
\usepackage{amssymb}
%\usepackage{amsthm}
%\usepackage{caption}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{epstopdf}
%\usepackage{graphics} % for pdf, bitmapped graphics files
%\usepackage{epsfig} % for postscript graphics files
\usepackage{mathptmx} % assumes new font selection scheme installed
%\usepackage{times} % assumes new font selection scheme installed
%\usepackage{amsmath} % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed
% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

@@

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Do not use the stfloats baselinefloat ability as IEEE does not allow
% \baselineskip to stretch. Authors submitting work to the IEEE should note
% that IEEE rarely uses double column equations and that authors should try
% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
% packages (also by Sigitas Tolusis) as IEEE does not format its papers in
% such ways.
% Do not attempt to use stfloats with fixltx2e as they are incompatible.
% Instead, use Morten Hogholm'a dblfloatfix which combines the features
% of both fixltx2e and stfloats:
%
% \usepackage{dblfloatfix}
% The latest version can be found at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/dblfloatfix/

@@

\begin{document}
% make the title area
%\begin{spacing}{2.5}
\date{}
\maketitle
\thispagestyle{empty}
\pagestyle{empty}
% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
Large MIMO constitutes a principal techniques for next generation wireless communication systems. Detection algorithms for large MIMO tend to require long simulation times on conventional computer systems, in order to improve simulation efficiency for the decoders of large MIMO systems, we propose to use General Purpose Graphic Processing Units (GPGPU). In this paper Fixed Complexity Sphere Decoder (FCSD) is implemented for the large scale MIMO system based on the dominant framework of GPGPU-Computed Unified Device Architecture (CUDA). The proposed CUDA-FCSD can achieve considerable speedup over conventional CPU implementation while reaching the same Bit Error Rate (BER) performance. Furthermore the proposed algorithm demonstrates great potential to accelerate large MIMO systems.

\textbf{ Keywords: Large MIMO, FCSD, GPGPU, CUDA}
\end{abstract}

@@

\section{Introduction}
The present challenge in the wireless communication area is the sharp contrast between the increasing demand of mobile network speed for the transmission of rich media content with better QoS (Quality of Service), and the shortage of radio frequency spectrum. Multiple Input Multiple Output (MIMO) is of immense research interest that can help to solve this problem owing to several advantages it can offer. The core idea of MIMO is to use multiple antennas that at the transmitting and receiving sides, that provide parallel data pipes in space resulting in spatial multiplexing and spatial diversity gains\cite{oestges2010mimo}.

Large scale MIMO (also named massive MIMO) systems employ tens to hundreds of low-power low-price antennas collocated at the base station and serving several tens of terminals at the same time\cite{rusek2013scaling}. This technology originated from multiuser MIMO (MU-MIMO) systems\cite{larsson2013massive}. Large scale MIMO can achieve the full potential of conventional MIMO systems, as well as provide additional reduction of latency and improvement of system robustness. Detector is one of the key component of MIMO systems,
 a well known detector for MIMO system is the Sphere Decoder (SD) that perform as well as a Maximum Likelihood detector (MLD) with a dramatic complexity reduction\cite{viterbo1999universal}\cite{hassibi2005sphere}. There are two major problems with the traditional SD decoder: complexity depending on Signal to Noise Ratio (SNR), and sequential nature, which are contradictory with the requirement of a constant data processing throughput and parallel implementation for practical system. The Fixed Complexity Sphere Decoder (FCSD) overcomes these two disadvantages of the SD decoder by searching the solution along a fixed number of independent paths\cite{barbero2008fixing}.

General purpose Graphic processing Units Computing (GPGPU) has become a common trend in industry and research in areas that need massive computations, such as computer vision\cite{fung2008using}, computer games\cite{blewitt2013applicability}, signal processing\cite{van2011accelerating}\cite{6671435} and finance\cite{grauer2013accelerating}. Compared to CPU, the GPU can naturally cope with computational intensive tasks, since most of the hardware resource are allocated for data processing and parallel computation, rather than caching and flow control\cite{cook2013cuda}. In recent years GPU computing has provided significant advantages in data processing speeds over CPU computations\cite{nvidia2008programming}. In 2011, the NVIDIA GPUs have become the fastest supercomputer in the world\cite{cook2013cuda},  Computed unified Device Architecture (CUDA) is a programming model for GPGPU that was proposed by NIVIDIA and has become the dominant framework for GPGPU programming. It is based on popular programming languages(C, C++, Java, Python, Fortran, Direct Compute), such that the application developers in different areas do not need to learn a new programming language or get to know the hardware details. CUDA employs the Single Instruction Multiple Threads (SIMT) programming model, enables a GPU to implement parallel data processing, and has become a parallel computing platform for wide community of different applications.

@@

The rest of the paper is organized as follows. Section \ref{programming model} briefly introduces CUDA, GPU architecture and the programming model. In section \ref{GPUFCSD}, we present details of the implementation of CUDA-FCSD. In section \ref{simulation}, the speedup and bit error rate (BER) performance of CUDA-FCSD are presented and analyzed. Finally conclusions are drawn in section \ref{conclusion}.    
% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.

@@

\section{CUDA Programming Model and GPU Architecture} \label{programming model} 
\subsection{Single Instruction Multiple Thread (SIMT) Model}
CUDA programming model is a C programming extension, with special defined syntax and structure. The serial code is executed on the host (CPU) side, and massive parallel data processing is performed on the device (GPU) side, this is defined as a heterogeneous programming model. A thread is the basic programming unit of CUDA, a block of cooperative threads form a thread block, the threads in one block can synchronize their behavior and communicate. Fig. \ref{figure1} simply illustrate CUDA memory hierachy model.
\begin{figure}[htb]
\centering
\includegraphics[width=0.45\textwidth]{CUDA_programming_hierachy.eps}
\caption{ CUDA 3D memory hierachy $5\times 5\times 5$ grid and each block contains $5\times 5\times 5$ threads
\small  The threads can be organized into one-dimensional, two-dimensional or three-dimensional blocks. Similarly, blocks can also be grouped into a one-dimensional, two-dimensional, or three-dimensional grid.
}
\label{figure1}
\end{figure}

Threads are scheduled, managed, created and executed in a group of 32 threads called warp, the programs executes on the device side are defined as $\mathit{kernel}$ function, when a $\mathit{kernel}$ function is invoked, the thread blocks are partitioned into warps, the threads in one warps executes the same instruction simultaneously, the different warps are scheduled by warp scheduler, when one warp is waiting for data loading the other warps begin to execute, this is called Single Instruction Multiple Thread (SIMT) model. This programming architecture is askin to Single Instruction Multiple Data (SIMD), however unlike the SIMD vectoring machines, SIMT can branch the behavior of each threads to achieve thread-level parallelism. The GPU use massive number of threads to hide the long latency of global memory instead of using large cache as CPU. In a word CPU cores are designed to minimize latency for each thread at a time, whereas GPUs are designed to handle a large number of concurrent, lightweight threads in order to maximize throughput.
\subsection{Mircroarchitecture of GPU} 
In Fig. \ref{figure2} we use the GPU-GeForce GTX 760 as an example to show the microarchitecture of  GPU. The GPU works as a coprocessor system, consisting of Stream Multiprocessors (SM), which perform the multiply-add arithmetic operation. Furthermore, each SM has the special functional units(SFU) that can perform more complex arithmetics such as trigonometric functions and reciprocal square root. the blocks are executed on SM.  There are five types of memory, register, shared memory, global memory, constant memory and texture memory. Registers are allocated dynamically and privately to threads, which has most rapid access speed.  The shared memory, also called on-chip memory, which are located on SM and can be shared among the threads in one block, it has low latency and limited capacity. Global memory, which is off-chip memory has large latency and large capacity. Another two types of the memory, constant memory and texture memory, are located off-chip, but cached. Therefore the access speed of these two type of memory space is much faster than that from global memory. The variables invoked in \textit{kernel} functions can be stored in either of the these five types of memories.

Because of the resource limitation of SM the total number of threads can be defined in one block is limited. On current CUDA supported GPUs, one block can contains up to 1024 threads, and all the blocks are executed on SM. Thus the total number of threads ($T_{total}$) we can make use of is $T_{total}=N_{SM}\times N_{threads/SM} $, $N_{SM}$ and $N_{threads/SM}$ denote the number of SM and number of threads per SM.  
\begin{figure}[htb]
\centering
\includegraphics[width=0.45\textwidth]{High_view_of_CUDA_GPU_microarchitecture.eps}
\caption{ High View of CUDA CPU Microarchitecture
The GeForce GTX 760 has 6 SM  and the maximum number of threads allowed on each SM is 2048, thus the total number of threads that can work in parallel is $2048\times 6=12288$.
}
\label{figure2}
\end{figure}   
% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals/conferences use top floats
% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
% command of the stfloats package.
\section{GPU Based Acceleration of FCSD}\label{GPUFCSD}
\subsection{MIMO System Model}\label{system}
% no \IEEEPARstart
%This demo file is intended to serve as a ``starter file''
%for IEEE conference papers produced under \LaTeX\ using
%IEEEtran.cls version 1.8 and later.
% You must have at least 2 lines in the paragraph with the drop letter
% (should never be an issue)
%I wish you the best of success.
%\hfill mds
%\hfill December 27, 2012

We consider a complex uncoded spatial multiplexing MIMO system with $N_r$ receive and $N_t$ transmit antennas, $N_{r}\geq N_{t}$, over a flat fading channel, corresponding to the discrete time model, the $\mathbf{y}\in\mathbb{C}^{N_{r}\times 1}$ is the received symbol vector can be written as:
\begin{equation}
\mathbf{y}=\mathbf{H}\mathbf{s}+\mathbf{n}   \label{formula 1}
\end{equation}
$\mathbf{s}\in \mathbb{C}^{N_{t}\times 1}$ is the transmitted symbol vector, with components that are mutually independent and taken from the signal constellation $\mathbb{O}$ (4-QAM, 16-QAM, 64-QAM), of size $M$. All the possible transmitted symbol vectors forms a $N_{t}$ complex signal vector space $\mathbb{O}^{N_{t}}$. $\mathbb{E}[\mathbf{s}\mathbf{s}^{H}]=\mathbf{I}_{N_t}E_{s}$, where $E_{s}$ denotes the symbol average energy, and $\mathbb{E}[\cdot]$ denotes the expectation operation. $\mathbf{H}\in \mathbb{C}^{N_{r}\times N_{t}}$ denotes the Rayleigh fading channel propagation matrix with independent identity distributed(i.i.d) circularly symmetric complex Gaussian zero mean components of unit variance, $\mathbf{n}\in \mathbb{C}^{N_{r}\times 1}$ is the additive white Gaussian noise (AWGN) samples with zero mean components and $\mathbb{E}[\mathbf{n}\mathbf{n}^{H}]=\mathbf{I}_{N_{r}}N_{0}$, $N_{0}$ denotes the average noise power spectrum density and $\frac{E_{s}}{N_{0}}$ denotes the signal to noise ratio (SNR).

Assume the receiver has the perfect channel state information (CSI), meaning that $ \mathbf{H}$ is known at the receiver, as well as the SNR. The task of MIMO decoder is to recover $\mathbf{s}$ based on $\mathbf{y}$ and $\mathbf{H}$.
\subsection{Fixed Complexity Sphere Decoder}
From (\ref{formula 1}), the maximum likelihood detector (MLD) for a MIMO system is specified by:
\begin{equation}
\mathbf{s}_{ML}=arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{y}-\mathbf{H}\mathbf{s}||^{2} \label{formula 2}
\end{equation}
We consider a MIMO system with  $N_{t}=N_{r}$. Performing the QR decomposition to the channel propagation matrix
\begin{equation}
 \mathbf{H}=\mathbf{Q}\mathbf{R}  \label{QR}
\end{equation}
where $\mathbf{Q}\in \mathbb{C}^{N_{t}\times N_{t}}$ is an unitary matrix and $\mathbf{R}\in \mathbb{C}^{N_{t}\times N_{t}}$, is an upper triangular matrix, the basing on \cite{golub2012matrix}, we can write (\ref{formula 2}) as:
\begin{eqnarray}
\nonumber
\mathbf{s}_{ML}&=&arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{Q}^{H}\mathbf{y}-\mathbf{Q}^{H}\mathbf{H}\mathbf{s}||^{2}\\
&=& arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{Q}^{H}\mathbf{y}-\mathbf{R}\mathbf{s}||^{2} \label{formula 3}
\end{eqnarray}
Consider the unconstrained estimation of $\mathbf{s}$, $\mathbf{\hat{s}}=\mathbf{G}\mathbf{y}$, $\mathbf{G}=(\mathbf{H}^{H}\mathbf{H})^{-1}\mathbf{H}^{H}$ based on (\ref{QR}), we have  
\begin{equation}
\mathbf{\hat{s}}=\mathbf{G}\mathbf{y}=(\mathbf{R}^{H}\mathbf{R})^{-1}\mathbf{R}^{H}\mathbf{Q}^{H}\mathbf{y}
=\mathbf{R}^{-1}\mathbf{Q}^{H}\mathbf{y}
\label{unconstrained estimation}
\end{equation} 
Therefore $\mathbf{Q}^{H}\mathbf{y}=\mathbf{R}\mathbf{\hat{s}}$, and (\ref{formula 3}) can be written as
\begin{equation}
\mathbf{s}_{ML}=arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{R}(\mathbf{\hat{s}}-\mathbf{s})||^{2} \label{formula 4}
\end{equation}

Just like SD the FCSD also performs depth first searching, at the first $\rho$ node levels\footnote{when $N_{r}=N_{t}$ the FCSD can achieve the same diversity as the MLD if $\rho\geq \sqrt{N_{t}}-1$ \cite{barbero2008fixing}} the FCSD search all the possible signal symbols in constellation $\mathbb{O}^{N_{t}}$, which is called Full Expansion(FE) stage, therefore for the symbols $\widetilde{s}_{N_{t}}, \widetilde{s}_{N_{t}-1}\dots \widetilde{s}_{N_{t}-\rho+1}$, the FCSD has $M$ branches expansion at each symbol node,  as shown in Fig. \ref{figure4}, there are $\rho=2$ FE node levels and each node level has $M=4$ branches. According to (\ref{formula 4}), we have: 
\begin{equation}
\hat{s}^{F}_{i}=arg\min_{s\in \mathbb{O}^{N_{t}}}\{\sum_{i=1}^{N_{t}}|r_{ii}(\hat{s}_{i}-s_{i})+\sum_{j=i+1}^{N_{t}}r_{ij}(\hat{s}_{j}-s_{j})|^{2}\}  \label{formula 5}
\end{equation} 
$i\in [1,2\dots N_{t}]$, where $r_{ij}$ denotes the component at the $i$th row and $j$th column of the upper triangular matrix $\mathbf{R}$, $\hat{s}_{i}^{F}$ denotes the FCSD solution symbol. Based on (\ref{formula 5}) at the remaining $N_{t}-\rho$ levels of symbols, the FCSD performs decision feed back, therefore there are only one branch expansion for each symbol node at this stage, this stage is called Single Expansion (SE). 
Thus the total number of the branches is fixed, the FCSD works as a constant number of multiple path tree searching algorithm.
